\bfseries\large Phosphatidylinositol 4,5-bisphosphate regulates cilium transition zone maturation in \textit{Drosophila melanogaster}

### 2018-02-26 latex/paper.tex
Various edits

Ciliogenesis begins with the assembly of TZ proteins at the tips of
the basal body.
This nascent TZ undergoes maturation, where the structure and protein
composition changes, allowing for anchoring of the TZ to the membrane and
establishment of a compartmentalized space bounded by the ciliary membrane
and the TZ.
In the Drosophila male germline, nascent TZs first assemble on basal bodies
during G2 phase in spermatocytes and anchor to the plasma membrane.
This is followed by establishment of cilium-membrane connections
through a poorly understood mechanism
involving microtubule rearrangements \citep{gottardo2013cilium}, and a
ciliary membrane that persists through meiosis \citep{riparbelli2012assembly}.
TZ maturation has been described in
\textit{Paramecium} \citep{aubusson2015transition}
\textit{Caenorrhabditis elegans} \citep{serwas2017centrioles} and
\textit{Drosophila melanogaster} \citep{gottardo2013cilium}
and is most readily observed by an increase in TZ length.
We expect that this process is analogous to the initial steps of TZ assembly
through sequential recruitment of TZ proteins and reorganization
in vertebrates.

Here, we show that the membrane lipid phosphatidylinositol 4,5-bisphosphate (\PIP{})
is essential for proper TZ maturation in developing male germ cells
in \textit{D. melanogaster}.
\PIP{} is relatively enriched within the plasma membrane and the base of cilia
in the male germline.
We show that expression of the bacterial PIP phosphatase SigD
or by inactivation of the PIP 5-kinase Skittles (Sktl) that localizes to cilia
induces longer than normal TZs without affecting TZ assembly.
These hyperelongated TZs show hallmarks of aberrant function similar to those
found in previously studied TZ mutants,
including loss of cilium-membrane associations.
We also find that the \textit{onion rings} allele of the
ciliopathy-associated \textit{EXOC8} gene encoding the human Exo84
decouples TZ hyperelongation from loss of cilium-membrane contacts.
Our results suggest that the plasma membrane lipid \PIP{} regulates TZ length,
maturation and function, thereby acting to support ciliogenesis.

\section{Results}
\subsection{\PIP{} is essential for transition zone maturation}
To investigate whether reduction of cellular \PIP{} affects ciliogenesis in the
Drosophila male germline,
we used transgenic flies expressing the Salmonella phosphoinositide phosphatase SigD
under the $\beta$2-tubulin promoter (referred to herein as \sigd{}),
which is expressed exclusively in spermatocytes
\citep{wei2008depletion, fabian2010phosphatidylinositol}.
We previously showed that \sigd{} induces male sterility
due in part to defects in flagellar biosynthesis \citep{wei2008depletion}.
To understand whether this was due to aberrant TZ function,
we examined the localization of fluorescently-labelled
centriolar/basal body protein
Ana1 (CEP295 homolog) \citep{goshima2007genes, blachon2009proximal}
and the TZ protein Cep290 \citep{basiri2014migrating}
during early steps of cilium assembly in spermatocytes in both
wild-type controls and \sigd{}.
As shown in Figure 1A,
while Cep290 on nascent TZs look similar in wild-type and \sigd{}, 
it grew to significantly longer lengths in \sigd{}
compared to wild-type controls.
Ana1 lengths were not affected, and we did not observe an inverse
correlation between Cep290 and Ana1 length.
This hyperelongation phenotype was not specific to Cep290;
all TZ markers examined showed hyperelongation, including
acetylated tubulin \citep{gottardo2013cilium},
Chibby \citep{enjolras2012drosophila} and
MKS \citep{slaats2015mks1, vieillard2016transition} proteins (Figure 1).
TZ hyperelongation was a highly penetrant phenotype (\textgreater 90\%)
and showed high intra-cyst correlation (\textgreater 0.95) suggestive
of a dosage-based response to a shared cellular factor,
presumably SigD.
Hyperelongated TZs persisted through meiosis.
Despite TZ hyperlongation, however, axonemes were able to
elongate post-meiosis (Figure 1).
As we showed previously, the ultrastructure of these axonemes
was frequently aberrant,
either lacking symmetry completely or containing triplet microtubules
instead of doublets \citep{wei2008depletion}.

\subsection{The type I PIP kinase Skittles localizes to centrioles and regulates TZ length}
While \PIP{} is the major substrate for SigD in eukaryotic cells \textit{in vivo}
\citep{terebiznik2002elimination, zhou2001salmonella, sengupta2013depletion},
SigD can also dephosphorylate other PIPs \textit{in vitro}
\citep{norris1998sopb}.
To address whether TZ hyperelongation observed in \sigd{} represented a
physiologically relevant phenotype in response to decreased \PIP{},
we attempted to rescue it by
co-expressing fluorescently-tagged full-length Sktl (Sktl) with SigD
under the same promoter.
We found that Sktl expression was able to suppress TZ elongation in a cilium-autonomous
manner (Figure 2A and B).
We also found that \textit{sktl} clones using Flp/FRT-based mitotic recombination
showed penetrant TZ hyperelongation (Figure 2C).

Since Sktl expression suppressed TZ hyperelongation,
we investigated whether Sktl could localize to centrioles or cilia in the Drosophila germline.
Similar to the human PIPKI$\gamma$ isoform \citep{xu2014pipkigamma},
we found that Sktl localizes to basal bodies in the Drosophila male germline.
As shown in Figure 2, Drosophila PIPKIs arose from the recent duplication of
the ancestral PIPKI gene, and are not orthologous to specific vertebrate PIPKI isoforms.
Sktl has diverged more than its paralog PIP5K59B, and seems
to be functionally related to PIPKI$\gamma$ in its roles at cilia.
However, unlike the human PIPKI$\gamma$, which functions to license TZ assembly
by promoting CP110 removal from basal bodies \citep{xu2016phosphatidylinositol},
our results suggest that Sktl functions in
regulating TZ length in the Drosophila male germline
but not its assembly.
Cp110 is dispensable for cilium formation in Drosophila \citep{franz2013cp110}
and we found its localization to be unaffected in \sigd{} (supplementary figure 1).

\subsection{Hyperelongated transition zones show signs of functional defects}
While shorter spermatocyte TZs have been reported in mutants of TZ proteins in Drosophila
\citep{vieillard2016transition, pratt2016drosophila},
no genetic manipulation has been reported to induce hyperelongated TZs.
We sought to examine whether TZ hyperelongation due to SigD expression
affected TZ function.
During germ cell elongation following meiosis, the TZ detaches from
the basal body and migrates along the growing axoneme, maintaining a ciliary compartment
at the distal-most ~5$\mu$m where tubulin is presumably incorporated into the axoneme.
We found that
TZs that assembled in \sigd{} cells are non-migratory despite the growth of the
axoneme and the cell membrane, as shown by Unc and Cep290 localization.
We previously described ``comet-shaped'' localization of Unc-GFP in these cells,
and we find that these hyperelongated Unc signals persist into cellular elongation
despite elongation of the axoneme.

%% reword this, paragraph is garbage
We previously described the presence of occasional triplet microtubules
in electron micrographs of axonemes from \sigd{} \citep{wei2008depletion}.
While centrioles and basal bodies contain microtubule triplets,
restriction of the C-tubule at the TZ causes the axoneme to
consist of microtubule doublets rather than triplets throughout its length.
Consistent with a defect in TZ barrier formation and presence of
microtubule triplets in axonemes,
we observed that a subset of cilia in \sigd{}
contained puncta of Ana1, a core centriolar protein, at the tips
of TZs marked with Cep290.
This TZ-distal Ana1 localization is observed at a relatively low penetrance
and high intra-cyst correlation.
We found that treatment of germ cells with Paclitaxel as in \citep{riparbelli2013unique}
could increase the penetrance of this phenotype.
Paclitaxel treatment stabilizes microtubules and is thought to disrupt TZ maturation
by interrupting microtubule rearrangements, and similarly induces much longer cilia.
Paclitaxel treatment alone did not induce this phenotype in wild-type controls in our hands
at varying concentrations ($p$ <0.01 with 5\% penetrance),
suggesting that loss of \PIP{} acts upstream of or prior to microtubule rearrangement.
Triplet microtubules within the axoneme are also observed in Taxol-treated cells \citep{riparbelli2013unique}.
It is possible that the TZ defect allows C-tubules and some centriolar proteins
to migrate with it to the distal end.
We did not observe this behaviour with Asterless,
a migratory protein that localizes to the outer
ring of centrioles ($p$ <0.01).

\subsection{The \textit{onion rings} mutant decouples defects found in cells with reduced levels of \PIP{}}
A heterozygous mutation in the human \textit{EXOC8},
the human ortholog of yeast Exo84, is linked to
Joubert syndrome \citep{dixon2012exome}.
Exo84 is a component of the octameric exocyst complex, which binds
\PIP{} and regulates membrane trafficking at the plasma membrane.
We previously showed that male flies homozygous for the \textit{onion rings}
(\textit{onr}) mutant of Drosophila Exo84 were sterile, with phenotypes
similar to \sigd{} \citep{wei2008depletion}.
To investigate whether defects in TZ hyperelongation could be explained by
defective Exo84 function, we examined TZs in \textit{onr} mutants.
We found that the \textit{onr} mutant did not display hyperelongated TZs (Figure 5),
suggesting that Exo84 is dispensable for TZ maturation.

Due to the involvement of exocyst in membrane trafficking at the plasma membrane,
we examined whether cilium-associated membrane dynamics were affected in \sigd{}
or \textit{onr} mutants similar to \textit{dila; cby} mutants
\citep{vieillard2016transition}.
We found that while TZs in \sigd{} and \textit{onr} cells anchored properly
to the plasma membrane, they were unable to either establish or maintain
membrane connections, and were rendered cytoplasmic (Figure 5) similar to
\textit{dila; cby} mutants.
Although we were not able to determine the localization of Exo84 due to the
lack of available antibodies,
we found that Sec6 and Exo70 localized to the plasma membrane as expected,
but were not enriched at the cilium (Figure 5C).
Our results suggest that the exocyst may act downstream of \PIP{} to
enhance cilium-membrane associations, and that TZ hyperelongation and loss of
membrane-cilium association are genetically separable phenotypes.

@@

In the Drosophila male germline, TZ maturation follows the initial
assembly of TZ proteins on the nascent basal body in the cytoplasm
followed by anchoring at the plasma membrane.
TZ maturation involves microtubule rearrangements and TZ elongation,
which leads to cilium-membrane attachment.
Loss of \PIP{}, a membrane lipid largely at the plasma membrane,
disrupts this process.
It causes excessive lengthening of the TZ,
and disrupts membrane-cilium connections.
Thus, intermediate steps during TZ maturation require \PIP{}.

>>>>

Ciliogenesis begins with the assembly of TZ proteins at the tips of
the basal body.
This nascent TZ undergoes maturation, during which its structure and protein
composition changes, allowing for anchoring of the TZ to the membrane and
establishment of a compartmentalized space bounded by the ciliary membrane
and the TZ.
In the Drosophila male germline, nascent TZs first assemble on basal bodies
during early G2 phase in spermatocytes and anchor to the plasma membrane.
This is followed by anchoring of cilia to the plasma membrane,
rearrangment of microtubules in the axoneme \citep{gottardo2013cilium},
and establishment of a
ciliary membrane that persists through meiosis \citep{riparbelli2012assembly}.
TZ maturation has been described in
\textit{Paramecium} \citep{aubusson2015transition}
\textit{Caenorrhabditis elegans} \citep{serwas2017centrioles} and
\textit{Drosophila melanogaster} \citep{gottardo2013cilium}
and is most readily observed by an increase in TZ length
in the Drosophila male germline.

Here, we show that the membrane lipid phosphatidylinositol 4,5-bisphosphate (\PIP{})
is essential for proper TZ maturation in developing male germ cells
in \textit{D. melanogaster}.
\PIP{} is relatively enriched within the plasma membrane and the base of cilia
in the male germline.
We show that expression of the bacterial PIP phosphatase SigD
or inactivation of the PIP 5-kinase Skittles (Sktl) that localizes to cilia
induces longer than normal TZs without affecting TZ assembly.
These hyperelongated TZs show hallmarks of aberrant function similar to those
found in previously studied TZ mutants,
including loss of cilium-membrane associations.
We also find that the \textit{onion rings} allele of the
ciliopathy-associated \textit{EXOC8} gene encoding the human Exo84
decouples TZ hyperelongation from loss of cilium-membrane contacts.
Our results suggest that the plasma membrane lipid \PIP{} regulates TZ length,
maturation and function, thereby acting to support ciliogenesis.

\section{Results}
\subsection{\PIP{} is essential for transition zone maturation}
To investigate whether reduction of cellular \PIP{} affects ciliogenesis in the
Drosophila male germline,
we used transgenic flies expressing the Salmonella phosphoinositide phosphatase SigD
under the $\beta$2-tubulin promoter (referred to herein as \sigd{}),
which is expressed exclusively in spermatocytes
\citep{wei2008depletion, fabian2010phosphatidylinositol}.
We previously showed that \sigd{} induces male sterility
due in part to defects in flagellar biosynthesis \citep{wei2008depletion}.
To examine whether this may be caused by aberrant TZ function,
we examined the localization of fluorescently-labelled
centriolar/basal body protein
Ana1 (CEP295 homolog) \citep{goshima2007genes, blachon2009proximal}
and the TZ protein Cep290 \citep{basiri2014migrating}
during early steps of cilium assembly in spermatocytes in
wild-type controls and \sigd{}.
As shown in Figure 1,
while Cep290 on nascent TZs look similar in wild-type and \sigd{}, 
it grew to significantly longer lengths in \sigd{}
compared to wild-type controls.
Ana1 lengths were not affected, and we did not observe an inverse
correlation between Cep290 and Ana1 length.
This hyperelongation phenotype was not specific to Cep290;
all TZ markers examined showed hyperelongation, including
acetylated tubulin \citep{gottardo2013cilium},
Chibby \citep{enjolras2012drosophila} and
MKS \citep{slaats2015mks1, vieillard2016transition} proteins (Figure 1D).
TZ hyperelongation was a highly penetrant phenotype (\textgreater 90\%)
and showed high intra-cyst correlation (\textgreater 0.95) suggestive
of a dosage-based response to a shared cellular factor,
presumably SigD.
Hyperelongated TZs persisted through meiosis.
Despite TZ hyperlongation, however, axonemes were able to
elongate post-meiosis (Figure 1E).
As we showed previously, the ultrastructure of these axonemes
was frequently aberrant,
either lacking symmetry completely or containing triplet microtubules
instead of doublets \citep{wei2008depletion}.

\subsection{The type I PIP kinase Skittles localizes to centrioles and regulates TZ length}
While \PIP{} is the major substrate for SigD in eukaryotic cells \textit{in vivo}
\citep{terebiznik2002elimination, zhou2001salmonella, sengupta2013depletion},
SigD can also dephosphorylate other PIPs \textit{in vitro}
\citep{norris1998sopb}.
To address whether TZ hyperelongation observed in \sigd{} represented a
physiologically relevant phenotype in response to decreased \PIP{},
we attempted to rescue it by
co-expressing fluorescently-tagged full-length Sktl (Sktl) with SigD
under the same promoter.
We found that Sktl expression was able to suppress TZ elongation
in a cilium-autonomous manner (Figure 2A and B).
We also found that \textit{sktl} clones using Flp/FRT-based mitotic recombination
showed TZ hyperelongation (Figure 2C).

Since Sktl expression suppressed TZ hyperelongation,
we investigated whether Sktl could localize to centrioles or cilia
in the Drosophila germline.
Similar to the human PIPKI$\gamma$ isoform \citep{xu2014pipkigamma},
we found that Sktl localizes to basal bodies in the Drosophila male germline.
As shown in Figure 2, Drosophila PIPKIs arose from the recent duplication of
the ancestral PIPKI gene,
and are not orthologous to specific vertebrate PIPKI isoforms.
Sktl has diverged more than its paralog PIP5K59B, and seems
to be functionally related to PIPKI$\gamma$ in its roles at cilia,
as well as the \textit{C. elegans} PPK-1 \citep{xu2014pipkigamma}.
However, unlike the human PIPKI$\gamma$, which functions to license TZ assembly
by promoting CP110 removal from basal bodies \citep{xu2016phosphatidylinositol},
our results suggest that Sktl functions in
regulating TZ length in the Drosophila male germline
but not its assembly.
Cp110 is dispensable for cilium formation in Drosophila \citep{franz2013cp110}
and we found its localization to be unaffected in \sigd{} (Supplementary figure 1).

\subsection{Hyperelongated transition zones exhibit signs of functional defects}
While shorter spermatocyte TZs have been reported in mutants of TZ proteins in Drosophila
\citep{vieillard2016transition, pratt2016drosophila},
no genetic manipulation has been shown to induce hyperelongated TZs.
We sought to examine whether TZ hyperelongation due to SigD expression
affected TZ function.
During germ cell elongation following meiosis, the TZ detaches from
the basal body and migrates along the growing axoneme, maintaining a ciliary compartment
at the distal-most {\raise.17ex\hbox{$\scriptstyle\sim$}}5$\mu$m where tubulin
is presumably incorporated into the axoneme.
We found that
TZs that assembled in \sigd{} cells are non-migratory despite the growth of the
axoneme and the cell membrane, as shown by Unc and Cep290 localization.
We previously described ``comet-shaped'' localization of Unc-GFP in these cells,
and we find that these hyperelongated Unc signals persist into cellular elongation
despite elongation of the axoneme (Figure 3).

We previously described the presence of occasional triplet microtubules
in electron micrographs of axonemes from \sigd{} \citep{wei2008depletion}.
While centrioles and basal bodies contain microtubule triplets,
restriction of the C-tubule at the TZ causes the axoneme to
consist of microtubule doublets rather than triplets throughout its length.
Consistent with a defect in TZ barrier formation and presence of
microtubule triplets in axonemes,
we observed that a subset of cilia in \sigd{}
contained puncta of Ana1, a core centriolar protein, at the tips
of TZs marked with Cep290.
This TZ-distal Ana1 localization is observed at a relatively low penetrance
and high intra-cyst correlation.
We found that treatment of germ cells with Taxol as in \citep{riparbelli2013unique}
could increase the penetrance of this phenotype.
Taxol treatment stabilizes microtubules and is thought to disrupt TZ maturation
by interrupting microtubule rearrangements, and similarly induces much longer cilia
\citep{riparbelli2012assembly}.
Taxol treatment alone did not induce this phenotype in wild-type controls in our hands
at varying concentrations ($p$ <0.01 with 5\% penetrance),
suggesting that loss of \PIP{} acts upstream of or prior to microtubule rearrangement.
Triplet microtubules within the axoneme are also observed in Taxol-treated cells \citep{riparbelli2013unique}.
It is possible that the TZ defect allows C-tubules and some centriolar proteins
to migrate to the distal end.
We did not observe this behaviour with Asterless (CEP152)
\citep{dzhindzhev2010asterless, blachon2008drosophila},
another centriolar protein ($p$ <0.01), suggesting that these TZ-distal
sites are not fully centriolar in protein composition.

\subsection{The \textit{onion rings} mutant decouples defects found in cells with reduced levels of \PIP{}}
We previously showed that male flies homozygous for the \textit{onion rings}
(\textit{onr}) mutant of Drosophila Exo84 were sterile, with phenotypes
similar to \sigd{} \citep{wei2008depletion}.
Exo84 is a component of the octameric exocyst complex, which binds
\PIP{} and regulates membrane trafficking at the plasma membrane.
To investigate whether defects in TZ hyperelongation could be explained by
defective Exo84 function, we examined TZs in \textit{onr} mutants.
We found that the \textit{onr} mutant did not display hyperelongated TZs (Figure 5),
suggesting that Exo84 is dispensable for TZ maturation.

Due to the involvement of exocyst in membrane trafficking at the plasma membrane,
we examined whether cilium-associated membrane dynamics were affected in \sigd{}
or \textit{onr} mutants similar to \textit{dila; cby} mutants
\citep{vieillard2016transition}.
We found that while TZs in \sigd{} and \textit{onr} cells anchored properly
to the plasma membrane, they were unable to either establish or maintain
membrane connections, and were rendered cytoplasmic (Figure 5) similar to
\textit{dila; cby} mutants.
Although we were not able to determine the localization of Exo84 due to the
lack of available reagents,
we found that Exo70 localized to centrioles (Figure 4D),
but not Sec6 (Supplementary figure 2).
Exo70 can directly bind to \PIP{} \citep{he2007exo70}.
Our results suggest that the exocyst may act downstream of \PIP{} to
enhance cilium-membrane associations, and that TZ hyperelongation and loss of
membrane-cilium association are genetically separable phenotypes.

@@

In the Drosophila male germline, TZ maturation follows the initial
assembly of a nascent TZ on the basal body in the cytoplasm
followed by anchoring at the plasma membrane.
TZ maturation involves microtubule rearrangements and TZ elongation,
which leads to cilium-plasma membrane attachment.
We found that loss of \PIP{}, a membrane lipid enriched at the plasma membrane,
affects this process of TZ maturation.
It causes excessive lengthening of the TZ,
and disrupts membrane-cilium connections.
